HO unification from object language to meta language

### 2024-02-25 paper/main.tex
Update aux relation

\noindent
Since \elpiIn{Pm} is an higher order unification variable with \elpiIn{x}
in its scope, the unification problem \ref{problem:pa} admits one solution:

@@

\noindent
After unifying the head of rule \ref{clause:r3a} with the goal Elpi runs
the premise \elpiIn{link Pm A P} that is in charge of bringing the
assignment for \elpiIn{Pm} (that has type \elpiIn{tm -> tm})
back to the domain of Coq terms:

@@

%
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t}\\
  \unifcomplete{\Ee}{\Ue}{t}
\end{gather}

@@

\paragraph{Term equality: \Eo vs. \Ee}
\paragraph{Term unification: \Uo vs. \Ue}
\paragraph{Term dereferencing: $\rho s$ vs. $\sigma t$}
TODO

@@

\begin{figure}
  \begin{elpicode}
    % congruence
    type ~(\Eo)~ tm -> tm -> o.
    app L1 ~\Eo~app L2 :- forall2 ~(\Eo)~ L1 L2.
    lam F1 ~\Eo~lam F2 :- pi x\ x ~\Eo~x => F1 x ~\Eo~F2 x.
    con X ~\Eo~con X.
    fov N ~\Eo~fov N.
    % eta
    lam F ==o T :- pi x\ beta T [x] (T' x), 
      x ==o x => F x ==o T' x.
    T ==o lam F :- pi x\ beta T [x] (T' x), 
      x ==o x => T' x ==o F x.
    % beta
    app [lam X | TL] ~\Eo~T :- beta (lam X) TL T', T' ~\Eo~T.
    T ~\Eo~app [lam X | TL] :- beta (lam X) TL T', T ~\Eo~T'.
  \end{elpicode}
  \caption{Equal predicate}
  \label{code:eq-pred}
  \Description[code:equal-pred]{code:equal-pred}
\end{figure}

Though, this equality relation is to be extended with $\eta\beta$ reduction
rules to mirror the theory of the OL. This mean
that new rules for those two redexes are added in the implementation of
\elpiIn{fo_equal}.

If \elpiIn{fo_equal} is conceived to manage equality between terms of the OL,
the same equality predicate in the ML behave slightly different. By the given
definition of the ML, the ML allows $\eta\beta$ congruence of terms, but, since
the node \elpiIn{app} and \elpiIn{abs} are constructor representing the
applications and the abstractions of the OL, these two reduction rules cannot
applied on them. We build therefore a predicate \elpiIn{equal} working for terms
in the ML which is implemented merely with the rules for the \elpiIn{fo_equal}
predicate.\todo{not true for uv} For example, if \elpiIn{fo_equal []
(abs x\ [c"f", c]) (c"f")} is true in the object language, \elpiIn{equal [] (abs
x\ [c"f", c]) (c"f")} produces a failure.

@@

Mathematically, we what to prove the following property:

\begin{coqcode}
  forall (t  u : term_ol) (s : subst_ol) 
         (t' u': term_ml) (s': subst_ml), 
    equal_ol t u s <->
      comp t t' /\ comp u u' /\ 
      unif_ml t' u' s' /\ links ... /\ decomp s' s.
\end{coqcode}

$$Math\ formula$$

In the following section we explain how we deal with term (de)compilation and
unification variable linking.

\section[Compilation: fo\_tm to tm]{Compilation}

>>>>

\noindent
Since \elpiIn{Pm} is an higher-order unification variable with \elpiIn{x}
in its scope, the unification problem \ref{problem:pa} admits one solution:

@@

\noindent
After unifying the head of rule \ref{clause:r3a} with the goal, Elpi runs
the premise <<\elpiIn{link Pm A P}>> that is in charge of bringing the
assignment for \elpiIn{Pm} (that has type \elpiIn{tm -> tm})
back to the domain of Coq terms:

@@

%
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}
\end{gather}

@@

\paragraph{Term dereferencing: $\rho s$ vs. $\sigma t$}

Since in our encoding we explicitly drag substitution mappings, we need to
dereference variables of the two languages. This is particularly useful to check
if two terms in the OL (resp. in the ML) are equal. The constructors
representing rigid terms, namely \elpiIn{app}, \elpiIn{lam} and \elpiIn{con},
make the dereferencing procedure to recurse over the their subterms. The code
below display this behavior for the dereferencing performed by the OL:

\begin{elpicode}
  type fo_deref fo_subst -> tm -> tm -> o.
  fo_deref S (app L1) (app L2) :- forall2 (fo_deref S) L1 L2.
  fo_deref S (lam F1) (lam F2) :- 
    pi x\ fo_deref S x x => fo_deref S (F1 x) (F2 x).
  fo_deref _ (con X) (con X).
\end{elpicode}

A very similar implementation for the dereferencing performed by the ML, given 
in the appendix, can be encoded respecting the signature:

\begin{elpicode}
  type ho_deref ho_subst -> tm -> tm -> o.
\end{elpicode}

More attention should be payed when the dereferencing encounters a subterm that
is unification variable. The following new rules should be added to the previous
code snipper to complete the dereferencing of variables of the OL:

\begin{elpicode}
  fo_deref S (fov N) R :- assigned? N S T, fo_deref S T R.
  fo_deref S (app [fov N|L]) R :- 
    assigned? N S T, beta T L R', fo_deref S R' R.
  fo_deref _ (fov N) (fov N).
\end{elpicode}

In particular, 1) if the variable \elpiIn{fov N} is assigned to a term
\elpiIn{T} in a substitution mapping \elpiIn{S}, we return the result of
dereferencing \elpiIn{T}; 2) if the current term is the application of a
variable \elpiIn{fov N} assigned to a list of terms \elpiIn{L} and \elpiIn{uv N}
is assigned to a term \elpiIn{T}, we return the result of dereferencing the beta
reduction of \elpiIn{T} to \elpiIn{L}. This is due to the semantics of the
unification variables within the OL; 3) finally, if \elpiIn{uv N} is not
assigned, then the same variable is returned.

A different reasoning is to be addressed to the variables of the ML. Firstly, a
meta variable cannot appear in the \elpiIn{app} node as the first element of the
list, we will explain why in \cref{sec:compilation}; secondly, variables in the
ML have a scope which should be taken into account. The following few lines of
code outline the rules of dereferencing for the ML:

\begin{elpicode}
  ho_deref H (hov N L) R :- assigned? N H T, 
    ho_beta T L R', ho_deref H R' R.
  ho_deref H (hov X L) (hov X L1) :- 
    forall2 (ho_deref H) L L1.
\end{elpicode}

\noindent
where, 1) if the variable \elpiIn{hov N L} has an assignment \elpiIn{T}, we
return the dereferencing of the beta reduction \elpiIn{T} to the scope
\elpiIn{L}.

\begin{figure}
  \begin{elpicode}
    type ~(\Eo)~ tm -> tm -> o.
    % congruence
    app L1 ~\Eo~app L2 :- forall2 ~(\Eo)~ L1 L2.
    lam F1 ~\Eo~lam F2 :- pi x\ x ~\Eo~x => F1 x ~\Eo~F2 x.
    con X ~\Eo~con X.
    fov N ~\Eo~fov N.
    % eta
    lam F ~\Eo~T :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta1}{\eta_1}$)~
      x ~\Eo~x => F x ~\Eo~T' x.
    T ~\Eo~lam F :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta2}{\eta_2}$)~
      x ~\Eo~x => T' x ~\Eo~F x.
    % beta
    app [lam X | TL] ~\Eo~T :- beta (lam X) TL T', T' ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_1}$)~
    T ~\Eo~app [lam X | TL] :- beta (lam X) TL T', T ~\Eo~T'. ~($\customlabel{clause:beta2}{\beta_2}$)~
  \end{elpicode}
  \caption{Equal predicate}
  \label{code:eq-pred}
  \Description[code:equal-pred]{code:equal-pred}
\end{figure}

\paragraph{Term equality: \Eo vs. \Ee} We can test if two terms are equal
following the equational theory of the language being considered. In
\cref{code:eq-pred} we provide an implementation of the \Eo predicate. The first
four rules check if the two terms are equal regarding the structure of the
current node, that is, two terms are equal if they have same head and if
recursively each subterm is two by two equal. Moreover, since the theory of the
OL accepts $\eta$- and $\beta$-equivalence, then we implement the corresponding
rules.

\noindent
The equality relation for the ML, accepts $\eta\beta$-equivalence between terms
of the ML. Recall that \elpiIn{abs x\ f x}, is a valid $\eta$ expansion of the
function \elpiIn{f} and that \elpiIn{lam x\ app[f, x]} is not that equivalent to
\elpiIn{f} at meta level. However, since we are interested in using the
unification procure of the ML, by \cref{prop:correct-ml}, we can use the
\Ue relation to test, when needed if two terms are equal in the ML.

\paragraph{Term unification: \Uo vs. \Ue}
TODO

@@

@@

In the following section we explain how we deal with term (de)compilation and
links between unification variables.

\section[Compilation: fo\_tm to tm]{Compilation}
\label{sec:compilation}
